Normalization Circularity in Population-Based FA Analysis

### 2011-06-08 ANTsTBSS/paper.tex
to jpg
Minor edits.

The seminal work of Basser et al. \cite{Basser1994a,Basser1994} established diffusion tensor imaging (DTI) as a viable investigatory MRI technique.  DTI's sensitivity to brain architecture \cite{Basser1996,Assaf2008} enables multiple uses in assessing neuro-structural differences in cross-population studies \cite{Kubicki2005,Arnone2008,Kantarci2010,Rametti2010} including subjects with traumatic brain injury (TBI) \cite{Caeyenberghs2010,Jiang2010,Warner2010}.  Two popular comprehensive software packages for assessing population FA differences include the SPM Matlab-based toolkit%
\footnote{
http://www.fil.ion.ucl.ac.uk/spm/
}
for voxel-based morphometry (VBM)
and the tract-based spatial statistics (TBSS) framework \cite{Smith2006}.%
\footnote{
http://www.fmrib.ox.ac.uk/fsl/tbss/index.html
} 
VBM \cite{Ashburner2001} continues to find application to FA studies \cite{Kakeda2010,Takao2010} despite concerns about its validity \cite{Bookstein2001,Davatzikos2004} in DTI \cite{Jones2005,Chung2008}.  The TBSS framework was partially developed in response to these concerns in which voxel values are projected onto a template white matter skeleton for increased statistical power.  
%(e.g. \cite{Arnone2008,Bodini2009}).

Although there are substantive differences in normalization schemes
between these and other image registration approaches, a core
commonality includes direct FA-to-FA registration.  The sum of squared differences (SSD) image similarity metric is perhaps the easiest to interpret (it drives the image difference to zero), is computationally efficient (just compute the image difference and an image gradient) and is therefore widely used.  Some of the most popular registration methods (SPM, Demons, FSL's FNIRT) rely on this metric and produce reasonable performance levels \cite{Klein2009}.%
\footnote{
In addition to TBSS and SPM, other popular packages in the neuroimaging community 
which use the SSD metric include AIR \cite{Woods1998} and LDDMM \cite{Beg2005}.
}

However, suppose the SSD metric is optimized to find a set of $M$ transformations that map a population of $M$ FA images, $\{I_1, I_2, \ldots, I_M\}$, to a representative FA template, $J$. The algorithm will then compute the transformations,
$\boldsymbol{\mathcal{T}} = \{\mathcal{T}_1, \mathcal{T}_2,\ldots,\mathcal{T}_M \}$, which minimize the SSD metric over the population, i.e.
\begin{align}
  \boldsymbol{\mathcal{T}} = \underset{\mathcal{T}_1, \ldots, \mathcal{T}_M}{\operatorname{argmin}}
    \sum_{m=1}^M\sum_{n=1}^N \left( I^n_m( \mathcal{T}_m(\mathbf{x}) ) - J \right)^2.
\end{align}
Switching the order of the summations (since the transformations are mutually independent) and recognizing that a principal criterion for the selection of $J$ is such that it be a good approximation of the mean of the aligned images, it becomes apparent that the transformation solution
\begin{align}\label{eq:variance}
  \boldsymbol{\mathcal{T}} = \underset{\mathcal{T}_1, \ldots, \mathcal{T}_M}{\operatorname{argmin}}
    \sum_{n=1}^N \underbrace{\sum_{m=1}^M \left( I^n_m( \mathcal{T}_m(\mathbf{x}) ) - J \right)^2}_{
    \propto\mathrm{\,voxelwise\,variance}}
\end{align}
is that which minimizes the voxelwise variance.  Reducing variance will also directly affect the value of a standard population statistic, the Student's $t$-test.  Other standard similarity metrics will also exhibit this
trend to some degree as they minimize some local measure of intensity-based distance between the subject and the template.

Thus, instead of normalization based on anatomical alignment followed by
statistical testing, direct FA normalization explicitly conflates
anatomical alignment with optimizing the very statistical testing
results that one will ultimately use to assess hypotheses.  This immediately evokes a sense of circularity in the analysis \cite{Kriegeskorte2010}.
According to Kriegeskrote et al., ``An analysis is circular (or nonindependent) if it is based on data that were selected for showing the effect of interest or a related effect.''  As illustrated with
Eqn. \ref{eq:variance}, direct FA-to-FA registration will find transformations that align (or select) voxels such that the statistical testing result (or the effect of interest) is increased (assuming one is using Student's 
$t$-test).  While the SSD metric is able to provide visually attractive alignments across subjects, the voxel-level details (that is, which voxels are matched where) are, as we show in the following simulation study, strikingly important in terms of the effect on detection power.

@@

\begin{figure}
\begin{tabular}{c}
  \includegraphics[width=85mm]{simulatedDataWarped.pdf}
\end{tabular}
\caption{Warping the image in Fig. \ref{fig:simulated_data}(c) to the template in 
  Fig. \ref{fig:simulated_data}(b) using the SSD similarity metric produces the image shown in
  (a).  The corresponding deformation is given in (b).  Although the images were anatomically
  aligned prior to invoking ANTs registration, deformation is driven by the similarity metric
  causing the accumulation of type 1 errors.}
\label{fig:simulated_data_warped}
\end{figure}

@@

After creation of the simulated data set, we performed voxelwise testing (both $t$-testing
and Wilcoxon rank-sum testing, one-tailed, $\alpha = 0.95$) with false discovery
rate correction within the white matter to get ``true'' cumulative distribution functions (CDFs).  The CDF vertical axis shows the cumulative number of voxels with corrected $p$-values at or below a given significance level (horizontal axis) and thus may be used as a surrogate measure for spatial sensitivity.  From here, we denote these graphs as "statistical significance curves."  These curves are plotted in black in Fig. \ref{fig:simulated_testing}.  To illustrate the 
effect of similarity metric on statistical significance, we performed ANTs registration \cite{Avants2011} of each subject to the template
(Fig. \ref{fig:simulated_data}(b)) using cross correlation (CC), mutual information (MI), 
Demons \cite{Thirion1998}, and sum of squared differences (SSD). A representative deformed image
using the SSD metric and its corresponding deformed grid is given in 
Fig. \ref{fig:simulated_data_warped}. 
The resulting statistical significance curves are plotted with the true significance curves in Fig. 
\ref{fig:simulated_testing}.

This example demonstrates the undesirable phenomenon previously described.  Even though these images were anatomically aligned prior to registration, the resulting deformations (falsely) increased detection power.  This effect can be explained in the context of Eqn. (\ref{eq:variance}) and by the fact that we added noise to the underlying distribution which allows the deformable mapping algorithms the freedom to transform the positions of the most "powerful" voxels to minimize variance across the population (and increase the $t$-test).  Note, however, that this effect is increased by the fact that noise was added to the data.  Decreasing the standard deviation of the noise causes an increase in separation between  the SSD and Demon's metric pair with the MI and CC metric pair with the latter pair closer to the the true significance curve. Increasing the noise yields the same ranking with a more even spread between significance curves.

@@

\begin{figure*}
\begin{center}
\begin{tabular}{c}
  \includegraphics[width=170mm]{workflow.pdf}
\end{tabular}
\caption{Possible ANTs-based TBSS workflow.  In contrast to standard TBSS in which an individual subject's FA is registered to the FMRIB58 template
(transform represented with the blue arrow), our proposed modification first constructs an ANTs optimal anatomical template as illustrated in the center of the figure.  Each FA image is then rigidly mapped, with distortion correction, to the individual T1 using the white matter mask as shown on the left.  The individual T1-weighted image is then nonrigidly registered to the anatomical template.  Optionally, one can register the T1 template to the MNI152 template which resides in the same space as the FMRIB58 template.  Thus, the composite transform, carefully constructed without any FA-to-FA image registration, can take an individual subject's FA map (or other DTI-derived measure) to the desired standardized coordinate system.    }
\label{fig:workflow}
\end{center}
\end{figure*}

@@

Population-specific templates (or optimal templates) alleviate some of
these issues by deriving a most representative image from the population
\cite{Good2001}.  Large deformation registration algorithms also reduce
this confound by being less sensitive to the deformation distance
between subject and target.  Some approaches combine both advantages,
for instance, the diffeomorphic approach of Joshi et al. who employ the
SSD metric and a shape distance to bring the subject group of images
into alignment \cite{Joshi2004}.  Variants
include extension to multiple modalities \cite{Lorenzen2006} and small deformations
\cite{Geng2009}.  These approaches iteratively minimize group difference in ``congealing''
towards a representative image template \cite{Learned-Miller2006}.

@@

\subsection{Anatomically-Based Population Normalization}
We now detail a standard approach that uses SyGN and intra-subject inter-modality mapping to provide an unbiased strategy for FA normalization.
An example of our proposed workflow applied to anatomically-based population normalization prior to FA analysis is illustrated in Fig. \ref{fig:workflow}.  Instead of direct FA-to-FA registration, our ANTs-based modification can be described as follows:

\begin{enumerate}
  \item Build the anatomical template from  the population sample or sample subset using SyGN as described in
  the previous section.
  \item If only using a sample subset for template creation, register the remaining
        anatomical data to the template and save the resulting transformation for each         
        subject indexed by $i$ (i.e. $\mathcal{T}^i_{template}$).
  \item Find the optimal transformation from each
        subject's FA to the anatomical template:
  \begin{enumerate}
  \item For each anatomical image, segment the white matter (in ANTs, one can use N4 bias correction \cite{Tustison2010} and Atropos $n$-tissue segmentation \cite{Avants2011a}).
  \item Rigidly transform each subject's average DWI to the bias-corrected anatomical image ($\mathcal{R}^i_{dwi}$).
  \item Perform distortion correction on the FA to anatomical mapping 
        by localized nonrigid registration
        of rigidly transformed FA map ($=\mathcal{R}^i_{dwi} \mathrm{FA}$) using
        the white matter mask obtained in step 3(a) ($\mathcal{T}^i_{FA}$).
  \item For each subject, concatenate the transforms which maps each
        subject's FA to the template ($\mathcal{T}^i_{total} = \mathcal{T}^i_{template} \circ 
        \mathcal{T}^i_{FA} \circ \mathcal{R}^i_{dwi} \mathrm{FA}$).
  \item Warp each subject's FA map to the template using $\mathcal{T}^i_{total}$.  Equivalently, one could warp the mean diffusion (MD), axial diffusion (AD) or any other scalar derivable from DTI to the template.  One other detail is that we prefer to warp the subject's DTI to the template and then, in template space, derive the scalar images of interest.  
  \end{enumerate}      
%  \item Perform the remaining portions of the standard TBSS pipeline:
%    \begin{enumerate}
%      \item Create the mean FA image from all template-aligned FA images.
%      \item Produce the white matter skeleton from the mean FA image.
%      \item Create skeleton distance map and project all FA data onto the skeleton.
%      \item Run the statistical analysis.
%    \end{enumerate}
\end{enumerate}
The next section compares this framework to FA-based normalization strategies with respect to both a patient-control study and a study of gender and age effects in a control population.

@@

In support of open science, the 1000 Functional Connectomes Project (FCP)%
\footnote{
ttp://fcon\_1000.projects.nitrc.org
} 
was initiated on December 11, 2009 by various members of the MRI community \cite{Biswal2010}.  Motivated by the absence of neuroimaging data for research, this initiative seeks to form collaborative partnerships with imaging institutions for sharing well-documented multi-modal image sets accompanied by phenotypic data.
Commitments from institutions such as they NYU Institute for Pediatric Neuroscience and the Nathan Kline Institute (NKI)--Rockland have already resulted in prospective data repositories
currently distributed to the public via the Neuroimaging Informatics Tools and Resources Clearinghouse (NITRC).%
\footnote{
http://www.nitrc.org
}
%Data from these two institutions were used to demonstrate the alignment properties obtained using SyGN for DTI-derived data normalization.

@@

The CDF of the voxelwise FA coefficient of variation of the aligned
images is shown for each case in Figure \ref{fig:tbi_cov}.  
The coefficient of variation at each voxel is defined as the variance
divided by the mean over the population of FA values.  Thus,
those metrics which best minimize the voxelwise variance would
trend towards greater alignment as quantified by the coefficient of
variation.

%This is very much related to earlier findings in that strategies evincing
%greater alignment, as quantified by the coefficient of variation,
%will trend towards smaller variance thus exhibiting a left-shifted
%CDF.

Although direct FA-to-FA registration using ANTs with common
similarity metrics gives ``better alignment'' such alignment is 
produced at a cost of significant type 1 errors.  Common subsampling and
smoothing strategies and local neighborhood metrics such as CC and MI reduce the 
Type 1 error problem (as evidenced in Fig. \ref{fig:tbi_testing}) 
but also show a decrease in alignment.  This is further supported
by the overlap measures \cite{Crum2006} given in Table \ref{table:lom}.
For each alignment strategy, we thresholded the derived mean FA image (FA $
\geq$ 0.2) and compared it with white matter mask of the MNI152 template
segmented using Atropos \cite{Avants2011a}.  The template-based
strategy showed the best overall overlap with the white matter MNI152
segmentation.

Finally, qualitative assessment is provided in Fig. \ref{fig:meanFA} 
in which the mean FA image derived from the template-based approach
and the post-normalization component of the TBSS pipeline (Fig. \ref{fig:meanFA}(c)) 
is juxtaposed 
with the mean FA image produced by the TBSS pipeline alone (Fig.\ref{fig:meanFA}(a)).
The skeleton masks were produced by thresholding the skeleton values at 
FA $\geq 0.2$ and it would seem that the improved alignment produces 
a more extensive skeletal network for greater localization testing of 
significant differences, particularly in the outer white matter regions
(compare Fig. \ref{fig:meanFA}(b) and Fig. \ref{fig:meanFA}(d)).
%Given that TBSS is intrinsically a statistical method for finding local 
%differences, improved alignment would presumably provide increased regional sensitivity to differences.

\begin{figure}
  \begin{center}
\begin{tabular}{c}
  \includegraphics[width=85mm]{tbiCoV.pdf}
\end{tabular}
\caption{Comparison of the different alignment strategies in normalizing the TBI data to the FMRIB58 template.  Assessment is quantitated using the voxelwise coefficient of variation of the FA.  TBSS and the 4 ANTs metric
results were all obtained using direct registration involving subject FA and the FMRIB58 template.  The curve denoted as ``antsT1'' was also obtained in FMRIB58 space with the following transformation composition: subject FA $\rightarrow$ subject T1 $\rightarrow$ group T1 template $\rightarrow$ 
common T1 template $\rightarrow$ MNI152 (FMRIB58) template. }
\label{fig:tbi_cov}
\end{center}
\end{figure}

@@

Four templates were generated from $\{5,10,20,30\}$ randomly selected subjects 
to see if the number of subjects comprising the template affected the alignment 
performance. In addition to the composite transforms
 to take each subject's FA or MD image to the population-specific anatomical template, we also registered
each ANTs T1 template to the MNI152 template.  
This permitted a
standard space for qualitative and quantitative comparison with TBSS-aligned
FA and MD images.  
After template construction, the remaining
images were registered to the resulting template and, ultimately, 
to the MNI152 (FMRIB58) template via the template-to-MNI152 transformation.

Sample axial slices for the FA FMRIB58-aligned mean images for the NKI data 
set of template sizes of 5 and 30 images are provided in Fig. \ref{fig:NKI}.  
We also include the same axial slice  of the FMRIB58 template in Fig. \ref{fig:NKI} 
(a) for comparison.  In both sets of sample slices, visual inspection would seemingly 
indicate better 
performance with the ANTs-based FA mean images compared with the standard 
TBSS-produced FA mean image.  Also, there is little discernible difference
 between the different ANTs results.  Alignment differences between the 
 approaches are particularly apparent in the peripheral white matter where 
 the anatomical structure in the ANTs template-based FA mean images is visually differentiable.

Relative performance is quantified in the plots produced in Fig. \ref{fig:NKIplots}. 
The mean FA image derived from the standard TBSS approach was thresholded at FA $\geq 0.2$. 
 This binary mask obtained in the NKI cohort provided the standard group mask for 
 fair comparisons for all evaluations.    For all FMRIB58-aligned images the coefficient of 
 variation (CoV) was calculated in a voxel-wise fashion within the mask.  Intuitively, 
 one would expect a smaller variance across better aligned sets of voxel values thus 
 producing a lower CoV for a given mean value.  By calculating the CDF of the set of CoV values within the mask, quantitative assessment of alignment performance can be visualized.  
% In addition to producing CDF plots for voxels within the entire brain mask, we separated the brain into inner and outer sections to investigate the alignment difference between these two regions.  This division was accomplished by applying a distance transform \cite{Maurer2003} to the FSL-provided MNI152 brain mask and thresholding the result such that the outer 30 mm form the outer mask and the subtracted binary image forms the inner mask.  Note that only those voxels that were also within the thresholded FA mask were considered in the inner/outer analysis.

There is a visible improvement in alignment for both data sets using the ANTs-based approach over individual subject FA registration to the FMRIB template.  There is little difference between the ANTs results with only a slight trend in improved performance with the larger template subject sample sizes.  
%This was the case for all three regional CoV calculations for both data sets.  Also, as expected, there is greater misalignment in the outer region versus the inner core of the white matter mask regardless of methodology.  However, the performance tally was the same in both the inner and outer regions as was the case with the entire white matter mask.% although the disparity in performance was much less for the NYU data set as the NKI data set most likely due to the differences in resolution of the original DTI.

@@

%\begin{figure*}
%\begin{tabular}{c}
%  \includegraphics[width=180mm]{NYUplots2.pdf}
%\end{tabular}
%\caption{{\bf NYU alignment assessment:}  To compare alignments between the  standard TBSS approach of using the FMRIB58 template versus using the different ANTs templates, we calculated the coefficient of variation at each voxel (for both FA and MD) and plotted the CDF over the range of values $[0,0.8]$ accumulated over all the voxels inside the thresholded (FA $\geq 0.2$) standard TBSS mean FA image (see Fig. \ref{fig:NKI}(b)).  In addition to the whole brain analysis (first column), the coefficient of variation was separately calculated over the inner and outer portions of the white matter (second and third columns, respectively).  The CDF plots for the ANTs templates are all very similar (although trending towards slight improvement with larger template sample size) and illustrate the general improvement in FA and MD alignment over registration of each FA image to the FMRIB58 template.
%}
%\label{fig:NYUplots}
%\end{figure*}

>>>>

The seminal work of Basser et al. \cite{Basser1994a,Basser1994} established diffusion tensor imaging (DTI) as a viable investigatory MRI technique.  DTI's sensitivity to brain architecture \cite{Basser1996,Assaf2008} enables multiple uses in assessing neuro-structural differences in cross-population studies \cite{Kubicki2005,Arnone2008,Kantarci2010,Rametti2010} including subjects with traumatic brain injury (TBI) \cite{Caeyenberghs2010,Jiang2010,Warner2010}.  Two popular comprehensive software packages for assessing population FA differences include the SPM Matlab-based toolkit%
\footnote{
http://www.fil.ion.ucl.ac.uk/spm/
}
for voxel-based morphometry (VBM)
and the tract-based spatial statistics (TBSS) framework \cite{Smith2006}.%
\footnote{
http://www.fmrib.ox.ac.uk/fsl/tbss/index.html
} 
VBM \cite{Ashburner2001} continues to find application to FA studies \cite{Kakeda2010,Takao2010} despite concerns about its general validity \cite{Bookstein2001,Davatzikos2004} and with respect to DTI \cite{Jones2005,Chung2008}.  The TBSS framework was partially developed in response to these concerns in which voxel values are projected onto a template white matter skeleton for increased statistical power.  
%(e.g. \cite{Arnone2008,Bodini2009}).

Although there are substantive differences in normalization
between these and other analysis approaches, a core
commonality includes direct FA-to-FA registration.  The sum of squared differences (SSD) image similarity metric is perhaps the easiest to interpret (it drives the image difference to zero), is computationally efficient (just compute the image difference and gradient) and is therefore widely used.  Some of the most popular registration methods (SPM, Demons, FSL's FNIRT) rely on this metric and yield reasonable performance levels \cite{Klein2009}.%
\footnote{
In addition to TBSS and SPM, other popular packages in the neuroimaging community 
which use the SSD metric include AIR \cite{Woods1998} and LDDMM \cite{Beg2005}.
}

However, suppose the SSD metric is used to find a set of $M$ transformations that map a population of $M$ FA images, $\{I_1, I_2, \ldots, I_M\}$, to a representative FA template, $J$, for statistical testing between groups. This results in the transformations,
$\boldsymbol{\mathcal{T}} = \{\mathcal{T}_1, \mathcal{T}_2,\ldots,\mathcal{T}_M \}$, which minimize the SSD metric over the population, i.e.
\begin{align}
  \boldsymbol{\mathcal{T}} = \underset{\mathcal{T}_1, \ldots, \mathcal{T}_M}{\operatorname{argmin}}
    \sum_{m=1}^M\sum_{n=1}^N \left( I^n_m( \mathcal{T}_m(\mathbf{x}) ) - J \right)^2.
\end{align}
Switching the order of the summations (since the transformations are mutually independent) and recognizing that a principal criterion for the selection of $J$ is such that it be a good approximation of the mean of the aligned images, it becomes apparent that the transformation solution
\begin{align}\label{eq:variance}
  \boldsymbol{\mathcal{T}} = \underset{\mathcal{T}_1, \ldots, \mathcal{T}_M}{\operatorname{argmin}}
    \sum_{n=1}^N \underbrace{\sum_{m=1}^M \left( I^n_m( \mathcal{T}_m(\mathbf{x}) ) - J \right)^2}_{
    \propto\mathrm{\,voxelwise\,variance}}
\end{align}
is that which minimizes the voxelwise variance.  Reducing variance will also directly affect the value of a standard population statistic, the Student's $t$-test.  Other standard similarity metrics will also exhibit this
trend to some degree as they minimize some local measure of intensity-based distance between the subject and the template.

Thus, instead of normalization based on anatomical alignment followed by
statistical testing, direct FA normalization explicitly conflates
anatomical alignment with optimizing the very statistical testing
results that one will ultimately use to assess hypotheses.  This immediately evokes a sense of circularity in the analysis \cite{Kriegeskorte2010}.
According to Kriegeskrote et al., ``An analysis is circular (or nonindependent) if it is based on data that were selected for showing the effect of interest or a related effect.''  As illustrated with
Eqn. \ref{eq:variance}, direct FA-to-FA registration will produce transformations that align (or select) voxels such that the statistical testing result (or the effect of interest) is increased (assuming one is using Student's 
$t$-test).  While the SSD metric is able to provide visually attractive alignments across subjects, the voxel-level details (that is, which voxels are matched where) are, as we show in the following simulation study, strikingly important in terms of the effect on detection power.

@@

\begin{figure}
\begin{tabular}{c}
  \includegraphics[width=85mm]{simulatedDataWarped.pdf}
\end{tabular}
\caption{Warping the image in Fig. \ref{fig:simulated_data}(c) to the template in 
  Fig. \ref{fig:simulated_data}(b) using the SSD similarity metric produces the image shown in
  (a).  The corresponding deformation is given in (b).  Although the images were anatomically
  aligned prior to invoking ANTs registration, deformation is driven by the similarity metric
  causing the accumulation of Type 1 errors.}
\label{fig:simulated_data_warped}
\end{figure}

@@

After creation of the simulated data set, we performed voxelwise testing (both $t$-testing
and Wilcoxon rank-sum testing, one-tailed, $\alpha = 0.95$) with false discovery
rate correction within the white matter to get ``true'' cumulative distribution functions (CDFs).  The CDF vertical axis shows the cumulative number of voxels with corrected $p$-values at or below a given significance level (horizontal axis) and thus may be used as a surrogate measure for spatial sensitivity.  We denote these graphs as "statistical significance curves."  
The true curves for the simulated experiment are plotted in black in Fig. \ref{fig:simulated_testing}.  
We note that the amount of smoothing increases the proportion of significant voxels beyond
the prescribed 10\%.

To illustrate the 
effect of similarity metric on statistical significance, we performed ANTs registration \cite{Avants2011} of each subject to the template
(Fig. \ref{fig:simulated_data}(b)) using cross correlation (CC), mutual information (MI), 
Demons \cite{Thirion1998}, and sum of squared differences (SSD). A representative deformed image
using the SSD metric and its corresponding deformed grid is given in 
Fig. \ref{fig:simulated_data_warped}. 
The resulting statistical significance curves are plotted with the true significance curves in Fig. 
\ref{fig:simulated_testing}.

This example demonstrates the undesirable phenomenon previously described.  Even though these images were anatomically aligned prior to registration, the resulting deformations (falsely) increased detection power.  This effect can be explained in the context of Eqn. (\ref{eq:variance}) and by the fact that we added noise to the underlying distribution which allows the deformable mapping algorithms the freedom to transform the positions of the most "powerful" voxels to minimize variance across the population (and increase the $t$-test statistics).  Whereas  decreasing the standard deviation of the noise causes an increase in separation between  the SSD and Demon's metric pair with the MI and CC metric pair with the latter pair closer to the the true significance curve, increasing the noise yields the same ranking with a more even spread between significance curves.

@@

\begin{figure*}
\begin{center}
\begin{tabular}{c}
  \includegraphics[width=170mm]{workflow.pdf}
\end{tabular}
\caption{Possible ANTs-based TBSS workflow using SyGN.  In contrast to standard TBSS in which an individual subject's FA is registered to the FMRIB58 template
(transform represented with the blue arrow), our proposed modification first constructs an ANTs optimal anatomical template as illustrated in the center of the figure.  Each FA image is then rigidly mapped, with distortion correction, to the individual T1 using the white matter mask as shown on the left.  The individual T1-weighted image is then nonrigidly registered to the anatomical template.  Optionally, one can register the T1 template to the MNI152 template which resides in the same space as the FMRIB58 template.  Thus, the composite transform, carefully constructed without any FA-to-FA image registration, can take an individual subject's FA map (or other DTI-derived measure) to the desired standardized coordinate system.    }
\label{fig:workflow}
\end{center}
\end{figure*}

@@

Population-specific templates (or optimal templates) alleviate some of
these issues by deriving a most representative image from the population
\cite{Good2001}.  Large deformation registration algorithms also reduce
this confound by being less sensitive to the deformation distance
between subject and target.  Some approaches combine both advantages,
for instance, the diffeomorphic approach of Joshi et al. employs the
SSD metric and a shape distance to bring the subject group of images
into alignment \cite{Joshi2004}.  Variants
include extension to multiple modalities \cite{Lorenzen2006} and small deformations
\cite{Geng2009}.  These approaches iteratively minimize group difference in ``congealing''
towards a representative image template \cite{Learned-Miller2006}.

@@

\subsection{Anatomically-Based Population Normalization}
We now detail a standard approach that uses SyGN and intra-subject inter-modality mapping to provide an unbiased strategy for FA normalization.
An example of our proposed workflow applied to anatomically-based population normalization prior to FA analysis using TBSS is illustrated in Fig. \ref{fig:workflow}.  Instead of direct FA-to-FA registration, our ANTs-based modification can be described as follows:

\begin{enumerate}
  \item Build the anatomical template from  the population sample or sample subset using SyGN as described in
  the previous section.
  \item If only using a sample subset for template creation, register the remaining
        anatomical data to the template and save the resulting transformation for each         
        subject indexed by $i$ (i.e. $\mathcal{T}^i_{template}$).
  \item Find the optimal transformation from each
        subject's FA to the anatomical template:
  \begin{enumerate}
  \item For each anatomical image, segment the white matter (in ANTs, one can use N4 bias correction \cite{Tustison2010} and Atropos $n$-tissue segmentation \cite{Avants2011a}).
  \item Rigidly transform each subject's average DWI to the bias-corrected anatomical image ($\mathcal{R}^i_{dwi}$).
  \item Perform distortion correction on the FA to anatomical mapping 
        by localized nonrigid registration
        of rigidly transformed FA map ($=\mathcal{R}^i_{dwi} \mathrm{FA}$) using
        the white matter mask obtained in step 3(a) ($\mathcal{T}^i_{FA}$).
  \item For each subject, concatenate the transforms which maps each
        subject's FA to the template ($\mathcal{T}^i_{total} = \mathcal{T}^i_{template} \circ 
        \mathcal{T}^i_{FA} \circ \mathcal{R}^i_{dwi} \mathrm{FA}$).
  \item Warp each subject's FA map to the template using $\mathcal{T}^i_{total}$.  
  Equivalently, one could warp the mean diffusion (MD), axial diffusion (AD) or any 
  other scalar derivable from DTI to the template.  
%  One other detail is that we prefer to warp the subject's DTI to 
%  the template and then, in template space, derive the scalar images of interest.  
  \end{enumerate}      
%  \item Perform the remaining portions of the standard TBSS pipeline:
%    \begin{enumerate}
%      \item Create the mean FA image from all template-aligned FA images.
%      \item Produce the white matter skeleton from the mean FA image.
%      \item Create skeleton distance map and project all FA data onto the skeleton.
%      \item Run the statistical analysis.
%    \end{enumerate}
\end{enumerate}
The next section compares this framework to FA-based normalization strategies with respect to both a patient-control study and a study of gender and age effects in a control population.

@@

In support of open science, the 1000 Functional Connectomes Project (FCP)%
\footnote{
ttp://fcon\_1000.projects.nitrc.org
} 
was initiated on December 11, 2009 by various members of the MRI community \cite{Biswal2010}.  Motivated by the absence of neuroimaging data for research, this initiative seeks to form collaborative partnerships with imaging institutions for sharing well-documented multi-modal image sets accompanied by phenotypic data.
Commitments from institutions such as the NYU Institute for Pediatric Neuroscience and the Nathan Kline Institute (NKI)--Rockland have already resulted in prospective data repositories
currently distributed to the public via the Neuroimaging Informatics Tools and Resources Clearinghouse (NITRC).%
\footnote{
http://www.nitrc.org
}
%Data from these two institutions were used to demonstrate the alignment properties obtained using SyGN for DTI-derived data normalization.

@@

The CDF of the voxelwise FA coefficient of variation (CV) of the aligned
images is shown for each case in Figure \ref{fig:tbi_cov}.  
The CV at each voxel is defined as the variance
divided by the mean over the population of FA values.  Thus,
those metrics which best minimize the voxelwise variance would
trend towards greater alignment as quantified by the CV.

%This is very much related to earlier findings in that strategies evincing
%greater alignment, as quantified by the CV,
%will trend towards smaller variance thus exhibiting a left-shifted
%CDF.

Although direct FA-to-FA registration using ANTs with common
similarity metrics gives ``better alignment'' such alignment is 
produced at a cost of significant Type 1 errors.  Common subsampling and
smoothing strategies and local neighborhood metrics such as CC and MI reduce the 
Type 1 error problem (as evidenced in Fig. \ref{fig:tbi_testing}) 
but also show a decrease in alignment.  This is further supported
by the overlap measures \cite{Crum2006} given in Table \ref{table:lom}.
For each alignment strategy, we thresholded the derived mean FA image (FA $
\geq$ 0.2) and compared it with white matter mask of the MNI152 template
segmented using Atropos \cite{Avants2011a}.  The template-based
strategy showed the best overall overlap with the white matter MNI152
segmentation.

Finally, qualitative assessment is provided in Fig. \ref{fig:meanFA} 
in which the mean FA image derived from the template-based approach
and the post-normalization component of the TBSS pipeline (Fig. \ref{fig:meanFA}(c)) 
is juxtaposed 
with the mean FA image produced by the TBSS pipeline alone (Fig.\ref{fig:meanFA}(a)).
The skeleton masks were produced by thresholding the skeleton values at 
FA $\geq 0.2$ and it would appear that the improved alignment produces 
a more extensive skeletal network for greater localization testing of 
significant differences, particularly in the outer white matter regions
(compare Fig. \ref{fig:meanFA}(b) and Fig. \ref{fig:meanFA}(d)).
%Given that TBSS is intrinsically a statistical method for finding local 
%differences, improved alignment would presumably provide increased regional sensitivity to differences.

\begin{figure}
  \begin{center}
\begin{tabular}{c}
  \includegraphics[width=85mm]{tbiCoV.pdf}
\end{tabular}
\caption{Comparison of the different alignment strategies in normalizing the TBI data to the FMRIB58 template.  Assessment is quantitated using the voxelwise CV of the FA.  TBSS and the 4 ANTs metric
results were all obtained using direct registration involving subject FA and the FMRIB58 template.  The curve denoted as ``antsT1'' was also obtained in FMRIB58 space with the following transformation composition: subject FA $\rightarrow$ subject T1 $\rightarrow$ group T1 template $\rightarrow$ 
common T1 template $\rightarrow$ MNI152 (FMRIB58) template. }
\label{fig:tbi_cov}
\end{center}
\end{figure}

@@

Four templates were generated from $\{5,10,20,30\}$ randomly selected subjects 
to see if the number of subjects comprising the template affected the alignment 
performance. In addition to the composite transforms
 to take each subject's FA or MD image to the population-specific anatomical template, we also registered
each ANTs T1 template to the MNI152 template.  
This permitted a
standard space for qualitative and quantitative comparison with TBSS-aligned
FA and MD images.  
After template construction, the remaining
images were registered to the resulting template and, ultimately, 
to the MNI152 (FMRIB58) template via the template-to-MNI152 transformations.

Sample axial slices for the FA FMRIB58-aligned mean images for the NKI data 
set of template sizes of 5 and 30 images are provided in Fig. \ref{fig:NKI}.  
We also include the same axial slice  of the FMRIB58 template in Fig. \ref{fig:NKI} 
(a) for comparison.  In both sets of sample slices, visual inspection would  
indicate better 
performance with the ANTs-based FA mean images compared with the standard 
TBSS-produced FA mean image.  Also, there is little discernible difference
 between the different ANTs results.  Alignment differences between the 
 approaches are particularly apparent in the peripheral white matter where 
 the anatomical structure in the ANTs template-based FA mean images is visually differentiable.

Relative performance is quantified in the plots produced in Fig. \ref{fig:NKIplots}. 
The mean FA image derived from the standard TBSS approach was thresholded at FA $\geq 0.2$. 
 This binary mask obtained in the NKI cohort provided the standard group mask for 
 fair comparisons for all evaluations.    For all FMRIB58-aligned images, the CV 
 was calculated in a voxelwise fashion within the mask.  Intuitively, 
 one would expect a smaller variance across better aligned sets of voxel values thus 
 producing a lower CV for a given mean value.  By calculating the CDF of the set of CV values within the mask, quantitative assessment of alignment performance can be visualized.  
% In addition to producing CDF plots for voxels within the entire brain mask, we separated the brain into inner and outer sections to investigate the alignment difference between these two regions.  This division was accomplished by applying a distance transform \cite{Maurer2003} to the FSL-provided MNI152 brain mask and thresholding the result such that the outer 30 mm form the outer mask and the subtracted binary image forms the inner mask.  Note that only those voxels that were also within the thresholded FA mask were considered in the inner/outer analysis.

There is a visible improvement in alignment for both data sets using the ANTs-based approach over individual subject FA registration to the FMRIB template.  There is little difference between the ANTs results with only a slight trend in improved performance with the larger template subject sample sizes.  
%This was the case for all three regional CV calculations for both data sets.  Also, as expected, there is greater misalignment in the outer region versus the inner core of the white matter mask regardless of methodology.  However, the performance tally was the same in both the inner and outer regions as was the case with the entire white matter mask.% although the disparity in performance was much less for the NYU data set as the NKI data set most likely due to the differences in resolution of the original DTI.

@@

%\begin{figure*}
%\begin{tabular}{c}
%  \includegraphics[width=180mm]{NYUplots2.pdf}
%\end{tabular}
%\caption{{\bf NYU alignment assessment:}  To compare alignments between the  standard TBSS approach of using the FMRIB58 template versus using the different ANTs templates, we calculated the CV at each voxel (for both FA and MD) and plotted the CDF over the range of values $[0,0.8]$ accumulated over all the voxels inside the thresholded (FA $\geq 0.2$) standard TBSS mean FA image (see Fig. \ref{fig:NKI}(b)).  In addition to the whole brain analysis (first column), the CV was separately calculated over the inner and outer portions of the white matter (second and third columns, respectively).  The CDF plots for the ANTs templates are all very similar (although trending towards slight improvement with larger template sample size) and illustrate the general improvement in FA and MD alignment over registration of each FA image to the FMRIB58 template.
%}
%\label{fig:NYUplots}
%\end{figure*}
